Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-20 paper/main.tex
add comparison between estimated and fitted growth_rates
rephrasing

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability.
% TODO citations
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction techniques.
% TODO citaitons
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the desired phenomenon which aligns with the observed biological
reality thus omitting to actually estimate the necessary parameters directly.
% TODO citations
This reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

@@

Mechanical properties of rod-shaped bacteria have been studied in great detail
experimentally~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
They are an interesting target system since their physical form provides more surface information
compared to spherical cells.
In addition, parameters such as growth rates have been found to vary across multiple
cells~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity on a
single-cell level.

Computational models for rod-shaped bacteria allow researchers to systematically analyze their
properties and infer emerging phenomena~\cite{Cho2007,Jnsson2005}.
Existing tools can describe growth, division and many effects such as colony formation and
expansion~\cite{Matyjaszkiewicz2017,Doumic2020} but they are limited in describing the
aforementioned plastic or elastic deformations of the rod.
Furthermore, the literature on how to estimate the parameters of such models directly from data is
very sparse.

@@

Visualization methods are an integral part of any modeling endevaour.
They allow researchers to obtain feedback quickly and improve iteration times when constructing new
models.
Furthermore, we utilize them in our more advanced parameter estimation methods (see
section~\ref{section:parameter-estimation}) where they enable us to effectively compare
two bacterial configurations across division events.
In the future, we are planning to build on the current state and generate realistic microscopic
images and time-series.
Together with the parameter estimation methods, they can then be used to provide large amounts of
mechanistic data for the training cell-segmentation and cell-tracking machine-learning algorithms.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Physical Interactions}
\label{subsection:mechanical-model-interactions}

When calculating forces acting between the cells, we can use a simplified model to circumvent the
numerically expensive integration over the complete length of the rod.
Given a vertex $\vec{x}_i$ on one cell, we calculate the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{w}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

@@

>>>>

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability.
% TODO citations
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction techniques.
% TODO citaitons
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the desired phenomenon which aligns with the observed biological
reality.
But this process does not actually estimate the necessary parameters directly
% TODO citations
and reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

@@

Mechanical properties of rod-shaped bacteria have been studied in great detail
experimentally~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
Rod-shaped Bacteria are an interesting target system since their physical form provides more surface
information compared to other forms such as spherical cells.
In addition, parameters such as growth rates have been found to vary across multiple
cells~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity on a
single-cell level.

Computational models for rod-shaped bacteria allow researchers to systematically analyze their
properties and infer emergent phenomena~\cite{Cho2007,Jnsson2005}.
Existing tools can describe growth, division and many effects such as colony formation and
expansion~\cite{Matyjaszkiewicz2017,Doumic2020} but they are limited in describing the
aforementioned plastic or elastic deformations of the rod.
Furthermore, the literature on how to estimate the parameters of such models directly from data is
very sparse.

@@

Visualization methods are an integral part of any modeling endevaour.
They allow researchers to obtain feedback quickly and improve iteration times when constructing new
models.
Furthermore, we utilize them in our more advanced parameter estimation methods (see
section~\ref{section:parameter-estimation}) where they enable us to effectively compare
two bacterial configurations across division events.
In the future, we are planning to build on the current state and generate realistic microscopic
images and time-series.
Together with the parameter estimation methods, they can then be used to provide large amounts of
mechanistic data for the training of cell-segmentation and cell-tracking machine-learning
algorithms.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Physical Interactions - Mie Potential}

When calculating forces which act between the cells, we can use a simplified model to circumvent the
numerically expensive integration over the complete length of the rod.
Given a vertex $\vec{x}_i$ on one cell, we calculate the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{w}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

@@

\begin{table}
    \centering
    \begin{tabularx}{\textwidth}{l c c c c c c}
        \toprule
        % \multicolumn{6}{Growth Rates}\\
        Method & Cell $0$ & Cell $1$ & Cell $2$ & Cell $3$ & Cell $4$ & Cell $5$\\
        \midrule
        \ac{abm} & 0.0011527990 & 0.0014106040 & 0.0018761827 & 0.0016834959 & 0.0036106023 & 0.0015209642\\
        Direct   & 0.0014057073 & 0.0010807540 & 0.0023177727 & 0.0015120675 & 0.0020488179 & 0.0016726263\\
        % 0.0014057073 & 0.0010807540 & 0.0023177727 & 0.0015120675 & 0.0020488179 & 0.0016726263\\
        % direct & 0.0009785162 & 0.0012641432 & 0.0021701728 & 0.0015695889 & 0.0036255661 & 0.0018419186\\
        \bottomrule
    \end{tabularx}
    \caption{TODO}
\end{table}
